---
title: "Introduction to R"
author: "Reiko Okamoto"
date: 2025-08-01
format: 
  html: 
    embed-resources: true
---

## 👀 Overview of R

-   R is a programming language and environment designed for statistical computing and data analysis
-   It provides a wide range of statistical and graphical techniques, making it a powerful tool for data exploration, manipulation, visualization, and modeling
-   Pros
    -   Free software (accessible to everyone)
    -   Large, active community contributing to code enhancements, bug fixes, and extensive collection of packages
-   Cons
    -   Can be slower and might not handle memory as efficiently as other languages

## 💻 Getting started with RStudio

-   An IDE (integrated development environment) is a software that brings together all the main tools you need for coding in one place
-   For R, the most popular IDE is RStudio

### What is the purpose of each pane?

-   Script editor: writing and editing R code
    -   ***Create a new R script***
    -   ***Write a basic mathematical operation***
-   Console: executing R code and displaying output
    -   ***Execute the code by selecting the code and clicking on the Run button***
    -   ***Execute the code using the keyboard shortcut (Ctrl/Cmd + Enter)***
-   Environment: displaying objects, variables, and their values
-   Files/Plots/Packages: managing files, displaying plots, handling packages

## 📝 Quarto document

-   R script is fine for writing and running code, but it's not very good at telling the story of your analysis
-   Quarto is a file format that allows you to combine your code, its results, and your explanations into one document
-   It supports multiple output formats (e.g., HTML, PDF, Word document)
-   ***Click the Render button to turn our notes and code into a polished document we can share***

## 1️⃣ Data types and variables

-   Data types describe what kind of information you're working with and what you can do with that information

### Data types commonly used in health data analysis

-   Character (text; enclosed in quotes)
-   Numeric
-   Logical (Boolean; `TRUE` or `FALSE`)
-   Date

#### *Assign a string to a variable*

```{r data-type-01}
# YOUR CODE HERE
```

### **What is a variable?**

-   Containers for storing data
-   Using lowercase letters and underscores for variable names is considered good practice
-   R is case-sensitive so `greeting` and `GREETING` are different variables
-   Pick variable names that describe what kind of information the variable holds

#### *Create a numeric variable*

```{r data-type-02}
# YOUR CODE HERE
```

#### *Create a logical variable*

```{r data-type-03}
# YOUR CODE HERE
```

#### *Practice exercises*

1.  Create a character variable with your name
2.  Create a numeric variable to store a year

```{r data-type-04}
# YOUR CODE HERE
```

-   In R, numbers are usually stored as **numeric** by default. Although you can specify that a number is an **integer** (a distinct data type representing whole numbers) by adding an `L` (like `2025L`), it's rare to need this in everyday coding.

## 🏢 Data structures

-   Data structures define how data is organized and stored

### Data structures commonly used in health data analysis

-   Atomic vector
-   Factor
-   Data frame
-   List

### Atomic vector

-   A collection of elements of the same type

#### *Create vectors*

```{r data-str-01}
# YOUR CODE HERE
```

#### *Check the length of the vector*

```{r data-str-02}
# YOUR CODE HERE
```

#### *Retrieve the first value of the vector*

```{r data-str-03}
# YOUR CODE HERE
```

-   In R, indexing (accessing elements by their position) starts at 1, not 0

#### *Retrieve the second and fourth values of the vector*

```{r data-str-04}
# YOUR CODE HERE
```

### Factor

-   Used for categorical variables
-   Unlike character vectors, factors have a set of pre-defined levels
-   Why is it useful?
    -   Makes modeling easier because many functions recognize factors as categorical variables automatically

    -   Gives you control over category order when plotting (instead of default alphabetical order)

#### *Create a factor variable*

```{r data-str-05}
# YOUR CODE HERE
```

-   `levels` lets you set the order of the categories; if you don't specify, R orders them alphabetically

### Data frames

-   A collection of vectors that all have the same length
-   Each column can contain a different data type, but inside one column, all the data needs to be the same type
-   Used to represent tabular data

#### *Create a data frame*

```{r data-str-06}
# YOUR CODE HERE
```

#### *Access a specific value by row and column indices*

```{r data-str-07}
# YOUR CODE HERE
```

-   We'll learn how to access specific columns and rows using `dplyr` later on. For more information on indexing and slicing data frames using base R, check out [this article](https://www.geeksforgeeks.org/r-language/indexing-and-slicing-data-frames-in-r/).

## ➕ Operators

### Arithmetic operators

-   Takes two objects and performs a calculation on them
-   E.g., `+`, `-`, `*`, `/`

#### *Calculate the remainder after diving 17 by 5*

```{r operators-02}
# YOUR CODE HERE
```

-   Useful when doing calculations that repeat in cycles (e.g., days of the week)

#### *Raise 3 to the power of 2*

```{r operators-03}
# YOUR CODE HERE
```

### Relational operators

-   Compares two objects and returns `TRUE` or `FALSE`

#### *Check that 10 is equal to 10*

```{r operators-04}
# YOUR CODE HERE
```

#### *Check that 5 is not equal to 7*

```{r operators-05}
# YOUR CODE HERE
```

#### *Check that 8 is greater than (or equal to) 5*

```{r operators-06}
# YOUR CODE HERE
```

-   Same principle applies to less than (or equal to)

### Logical operators

-   Take in values that are (or can be turned into) `TRUE` or `FALSE` and returns `TRUE` or `FALSE`

#### *Check that 5 is greater than 3 and that 7 is less than 10*

```{r operators-07}
# YOUR CODE HERE
```

#### *Check that 4 is greater than 6 or that 9 is equal to 9*

```{r operators-08}
# YOUR CODE HERE
```

#### *Flip a Boolean value*

```{r operators-09}
# YOUR CODE HERE
```

-   Relational and logical operators are especially useful for filtering data in a data frame
-   Check out [this article](https://www.geeksforgeeks.org/r-language/r-operators/) for a comprehensive list of operators in R

## 🧮 Functions

-   Functions are blocks of code designed to perform specific tasks
-   They take inputs, process them, and return outputs

#### *Calculate the mean (average) of a numeric vector*

```{r functions-01}
# YOUR CODE HERE
```

#### *Calculate the sum of values in a numeric vector*

```{r functions-02}
# YOUR CODE HERE
```

#### *Round 3.14159 to 2 decimal places*

```{r functions-03}
# YOUR CODE HERE
```

-   The `help()` function or `?` can be used to access function documentation

#### *Other commonly used functions*

-   `median()` computes the median of a numeric vector
-   `min()` finds the smallest value in a numeric vector
-   `max()` finds the largest value in a numeric vector
-   `sd()` computes the standard deviation of a numeric vector
-   `abs()` computes the absolute value of a numeric vector
-   `sqrt()` computes the square root of a numeric vector

## 🧑‍🔧 Pipes

#### *Compute the sum of every number from 1 to 10, then take its square root, rounded to 3 decimal places*

```{r pipes-01}
# YOUR CODE HERE
```

#### *Repeat the calculation with the pipe operator*

```{r pipes-02}
# YOUR CODE HERE
```

-   The **pipe operator** (Ctrl/Cmd + Shift + M) allows users to chain operations together, enhancing code readability and reducing nesting
-   The operations flow left-to-right so the sequence of actions is presented more clearly
-   Anything written after `#` on a line is considered a **comment** and is not executed as code
    -   These help clarify the purpose or functionality of code for other users and even for yourself in the future

## 📦 Packages

-   **Packages** are sets of R functions, data sets, and other resources bundled together
-   They enhance base R's functionality by providing specialized tools for various tasks
-   We will focus on:
    -   [`dplyr`](https://github.com/tidyverse/dplyr) for data manipulation and transformation
    -   [`ggplot2`](https://github.com/tidyverse/ggplot2) for creating high-quality visualizations
-   Both of these packages belong to the [tidyverse](https://www.tidyverse.org/), a meta-package (package of packages) designed for data science and analysis

## 🏎️ If we have time: Test drive `dplyr`

#### *Load the package*

```{r dplyr-01}
# YOUR CODE HERE
```

#### *Filter rows where `mpg` (fuel efficiency) is greater than 20*

```{r dplyr-02}
# YOUR CODE HERE
```

#### *Select `mpg` and `cyl` columns*

```{r dplyr-03}
# YOUR CODE HERE
```

#### *Compute the power-to-weight ratio of each car*

```{r dplyr-04}
# YOUR CODE HERE
```

#### *Compute the average fuel efficiency*

```{r dplyr-05}
# YOUR CODE HERE
```

#### *Reorder the data by fuel efficiency in ascending order*

```{r dplyr-06}
# YOUR CODE HERE
```

#### *More practice exercises*

-   Using the `mtcars` data set and the pipe, perform the following steps in sequence:
    1.  Select the columns `mpg`, `cyl`, `hp`, and `wt`

    2.  Filter the rows to include only cars with `mpg` greater than 15

    3.  Create a new column `km_per_l` by converting `mpg` to kilometers per liter (1 mpg = 0.425144 km/L)

    4.  Calculate the average `kl_per_l`

```{r dplyr-07}
# YOUR CODE HERE
```

## 📚 Resources

-   [R for Data Science](https://r4ds.had.co.nz/)
